Metagenomic Barcoding of Pollen Loads Offers Insights on the Foraging

### 2023-01-13 script/manuscript_MEE.Rmd
some writing on management actions

To detect whether the sequencing reads were semi-quantitative the subset of all pollen morphotypes distinguishable by microscopy were compared to the sequence reads. In all instances sequence reads were pooled to the highest taxonomic rank associated with the morphotype, e.g. if both species of *Mertensia*, or one species and read only classified to genus were present in a sample, the reads were summed. The total percentage of the ten most abundant grains per sample were then were then relativized to constitute the entire sample.

The relationship shows that grains which are often present in trace amounts have relatively high amounts of their abundances estimated, and grains which are present in high amounts have relatively low amounts of sequences classified. This is likely due to the proportion of high false positives which occur in the classification process with NGS (BELL NOVEMEBER 2021). There was strong evidence of a strong correlation between the proportion of grains per morphotype and the number of sequences per group (`r round(ct[["estimate"]][["tau"]], 3)`, p < 0.0001, n = `r nrow(pol_p)`).

@@

We have demonstrated how Angiosperms533 hyb-seq probes may be used for plant barcoding in a metagenomic context. This was exemplified in an ecologically relevant scenario, where the results have immediate implications for natural history driven fundamental science and the applied science of land management. The test pollen loads contained a number of closely related taxa, some in notoriously taxonomically difficult clades (e.g. *Mertensia*, *Lupinus*), at naturally occurring proportions.  We incorporated spatial and temporal approaches for creating custom sequence databases an approach which is readily applicable to any lab group with the capacity to perform next-generation sequencing across the entirety of multiple continents, and which we expect to be highly beneficial in many study areas. By combining insights from these novel approaches with an extensive observational field based study we show how these methods may be applied to test a variety of hypotheses related to ecological interactions.

@@

**Spatial ... & Timing ** (filters!)
  We have concerns regarding the number of persons training to become and practice botany, and grave concerns regarding the funding mechanisms for floristic and field based botanical research and for centralized authorities to produce consensus opinions on alpha taxonomy (@prather2004decline, @kramer2015report, @prather2004implications, @crisci2020end, @manzano2021flippant, @stroud2022botanical). To reduce the effects of a low population density of botanists on the maintenance of and production of flora's and to foster metagenomics across landscapes without field stations we utilized Species Distribution Modelling to generate predictive species lists. In this proof of concept example we performed several iterations of modelling runs, and several approaches (i.e. the 'linear models', and the 'machine learning'), which took notable amounts of compute power. We suspect the possible deleterious nature of this endeavor may be reduced by: 1) more field surveying by crews will reduce the need to generate as many species 2) fewer runs of models, 3) only running machine learning models which do not require an explicitly process to reduce spatial autocorrelation. However, given the time required to perform all aspects of a study, even our amount of computation was negligible. Further, we are very optimistic about the possibility for persons to perform these tasks, as mentioned we utilized roughly only one quarter of the records which were digitally available for presence, and we suspect others will have enough records to perform this process nearly anywhere else in the temperate.
  Tandem to the lack of continued expertise required to generate and maintain species lists, is the expertise required to continue tracking when major phenological events occur in many plant species at relatively fine scales or under novel climates. Knowledge of these events is currently limited to general time periods of only a handful of phenological events and groups of organisms (e.g. flowering initiation, or trees) (@prather2004implications, @li2016responses). While many programs and initiatives exist to collect phenological information on subsets of easily identifiable charismatic species to detect major trends in phenology, by design these capture only a subset of the extent diversity (@betancourt2005implementing, @havens2007chicago). In many instances it appears that while landscapes respond similarly to environmental variables which predict phenological responses, that individual species vary widely in their responses to similar environmental cues, or respond to different cues (@augspurger2020concordance, @xie2015deciduous, @xie2018predicting, @caradonna2014shifts). As can be seen here, predictions of when a single, major phenological event occurs is already data limited, with sample size having an effect on the subset of species which we could even generate weibull estimates for.

**Molecular ... **
  The nearly complete Plant and Fungal Tree of Life (PAFTOL) will provide a comprehensive phylogenetic backbone of the entire plant kingdom, and the inclusion of A353 probes with lineage specific probe sets is common in producing massive genetic datasets (@baker2021exploring).  We predict that the A353 probes which it is utilizing to work nearly immediately for DNA barcoding of whole plant material, and that more elaborate validation studies in controlled metabarcoding settings, utilizing existing experimental designs, will have favorable results (@bell2017applying, @bell2019quantitative, @bell2021comparing, @lamb2019quantitative). In particular the harvesting of loci with more variation in certain lineages, and or with more variable flanking regions, will prove promising for identifying closely related plant material (CITE). We suspect that conserved reaches of genes resulted in the high amounts of reads in somewhat obscure species. Given that the A353 loci are nuclear, single copy, and a variety are present the possibility of identifying target loci for quantitative purposes is high, without continual PCR enrichment is possible; this would align with relatively high efficacy of WGS (@lang2019genome, @peel2019semi, @bell2021comparing). Recent evidence indicates that the potential for identifying nearly cryptic taxa and even infra-specific inference, of either whole plant material, and perhaps in metagenomic context are possible (@ottenlips2021resolving, @wenzell2021incomplete, Loke et al. in prep,  @slimp2021potential, @beck2021palmer).

# 4 | CONCLUSION 
We believe that the combination of spatial and temporal models, united and guided by localized natural history knowledge, provides the essential components of a bayesian framework for approaching the coarse elucidation of ecological interactions using DNA Barcoding. Herein we crudely utilized this thinking via binary outcomes, should a species predicted be predicted present or not? Is it unequivocally flowering or not? Myriad data show biological systems and ecological interactions have more variance than can be reasonably discretely parsed. We expect that within a bayesian framework studies of pollinator behavior may be enacted via this approach at a landscape level, e.g. the scale of an entire drainage basin such as the Gunnison which is quickly becoming one of the worlds few model ecosystems. We hope that the promise of A353 probes as tools for metabarcoding play a role in these endeavors.

>>>>

To detect whether the sequencing reads were semi-quantitative the proportion of all pollen morphotypes distinguishable by microscopy were compared to the sequence reads. In all instances sequence reads were pooled to the highest taxonomic rank associated with the morphotype, e.g. if both species of *Mertensia*, or one species and read only classified to genus were present in a sample, the reads were summed. The total percentage of the ten most abundant grains per sample were then were then *corrected* to constitute the entire sample.

The relationship between the number of pollen grains in a sample and the number of sequence reads is roughly *curvilinear*, where grains which are present in trace amounts are overestimated by sequence counts, while grains present in high amounts are underestimated. This is likely due to the proportion of high false positives which occur in the classification process with NGS (BELL NOVEMEBER 2021). There was strong evidence of a strong correlation between the proportion of grains per morphotype and the number of sequences per group (`r round(ct[["estimate"]][["tau"]], 3)`, p < 0.0001, n = `r nrow(pol_p)`).

@@

We have demonstrated how Angiosperms533 hyb-seq probes may be used for plant barcoding in a metagenomic context. This was exemplified in an ecologically relevant scenario, where the results have immediate implications for natural history driven fundamental science and the applied science of land management. The test pollen loads contained a number of closely related taxa, some in notoriously morphologically difficult clades with rapid rates of diversification (e.g. *Mertensia*, *Lupinus*), at naturally occurring proportions (@nevado2016widespread, @nazaire2014phylogenetic).  We incorporated spatial and temporal approaches for creating custom sequence databases an approach which is readily applicable to any lab group with the capacity to perform next-generation sequencing across the entirety of multiple continents, and which we expect to be highly beneficial in many study areas. By combining insights from these novel approaches with an extensive observational field based study we show how these methods may be applied to test a variety of hypotheses related to ecological interactions.

@@

  We have concerns regarding the number of persons training to become and practice botany, and grave concerns regarding the funding mechanisms for floristic and field based botanical research and for centralized authorities to produce consensus opinions on alpha taxonomy (@prather2004decline, @kramer2015report, @prather2004implications, @crisci2020end, @manzano2021flippant, @stroud2022botanical). To reduce the effects of a low population density of botanists on the maintenance of and production of flora's and to foster meta-genomics across landscapes without field stations we utilized Species Distribution Modelling to generate predictive species lists. In this proof of concept example we performed several iterations of modelling runs, and several approaches (i.e. the 'linear models', and the 'machine learning'), which took notable amounts of compute power. We suspect the possible deleterious nature of this endeavor may be reduced by: 1) more field surveying by crews will reduce the need to generate as many species 2) fewer runs of models, 3) only running machine learning models which do not require an explicitly process to reduce spatial autocorrelation. However, given the time required to perform all aspects of a study, even our amount of computation was negligible. Further, we are very optimistic about the possibility for persons to perform these tasks, as mentioned we utilized roughly only one quarter of the records which were digitally available for presence, and we suspect others will have enough records to perform this process nearly anywhere else in the temperate.
  Tandem to the lack of continued expertise required to generate and maintain species lists, is the expertise required to continue tracking when major phenological events occur in many plant species at relatively fine scales or under novel climates. Knowledge of these events is currently limited to general time periods of only a handful of phenological events and groups of organisms (e.g. flowering initiation, or trees) (@prather2004implications, @li2016responses). While many programs and initiatives exist to collect phenological information on subsets of easily identifiable charismatic species to detect major trends in phenology, by design these capture only a subset of the extent diversity (@betancourt2005implementing, @havens2007chicago). In many instances it appears that while landscapes respond similarly to environmental variables which predict phenological responses, that individual species vary widely in their responses to similar environmental cues, or respond to different cues (@augspurger2020concordance, @xie2015deciduous, @xie2018predicting, @caradonna2014shifts). As can be seen here, predictions of when a single, major phenological event occurs is already data limited, with sample size having an effect on the subset of species which we could even generate weibull estimates for.

  The nearly complete Plant and Fungal Tree of Life (PAFTOL) will provide a comprehensive phylogenetic backbone of the entire plant kingdom, and the inclusion of A353 probes with lineage specific probe sets is common in producing massive genetic datasets (@baker2021exploring).  We predict that the A353 probes which it is utilizing to work nearly immediately for DNA barcoding of whole plant material, and that more elaborate validation studies in controlled metabarcoding settings, utilizing existing experimental designs, will have favorable results (@bell2017applying, @bell2019quantitative, @bell2021comparing, @lamb2019quantitative). In particular the harvesting of loci with more variation in certain lineages, and or with more variable flanking regions, will prove promising for identifying closely related plant material (CITE). We suspect that conserved reaches of genes resulted in the high amounts of reads in somewhat obscure species. Given that the A353 loci are nuclear, single copy, and a variety are present the possibility of identifying target loci for quantitative purposes is high, without continual PCR enrichment is possible; this would align with relatively high efficacy of WGS (@lang2019genome, @peel2019semi, @bell2021comparing). Recent evidence indicates that the potential for identifying nearly cryptic taxa and even infra-specific inference, of either whole plant material, and perhaps in metagenomic context are possible (@ottenlips2021resolving, @wenzell2021incomplete, Loke et al. in prep,  @slimp2021potential, @beck2021palmer).

  The remarkably clear foraging preferences of *Bombus*, both here and across a great many localities globally, reiterates the needs for land managers to maintain relatively high amounts of members of the Fabaceae, Boraginaceae, and Ranunculaceae across the landscape. Numerous historic, and some ongoing, land management actions reduce the ability of many landscapes to support stable populations of *Bombus*. Historic, and occasionally current, livestock grazing may be associated with the targeted removal of many species of plants which are known to have compounds toxic to cattle. In particular, the removal of locoweeds (Fabaceae: *Astragalus* & *Oxytropis*) and larkspurs (Ranunculaeae: *Delphinium*) were common across public lands administered by the United States Forest Service, and to an extent the Bureau of Land Management (@ralphs1988herbicide, @aldous1919eradicating, @ralphs2003mechanism). Further actions, generally initiated by settlers to gain water rights or to support mining operations, involved the channelization and incising of streams, and culling of beavers, reducing the extent of wetlands and associated mesic meadows which provide habitat for many species of *Mertensia* (Boraginaceae) widely distributed across Western North America (). Fire suppression further resulted in the succession of many Aspen groves to Conifer stands, decreasing the mosaic of age structured habitats in many landscapes, adversely effection habtitat for tall *Mertensia* species. Finally the effects of Nitrogen deposition, especially given the West's growing population still pose adverse effects on the abundance of a variety of species of Fabaceae (see @stevens2018atmospheric). Current solutions to these issues, involve targeted burns, reintroduction of beavers and beaver habitats, and the possibility of re-seeding a variety of 'locoweeds' and 'larkspurs' in areas now seldom used for grazing. The highly enthusiastic response of land managers, and homeowners, to plant *Ascelpias* to improve Monarch Butterfly (*Danaus plexippus*) habitat provides an effective framework for the latter.

# 5 | CONCLUSION 
We believe that the combination of spatial and temporal models, united and guided by localized natural history knowledge, provides the essential components of a bayesian framework for approaching the coarse elucidation of ecological interactions using DNA Barcoding. Herein we crudely utilized this thinking via binary outcomes, should a species predicted be predicted present or not? Is it unequivocally flowering or not? Myriad data show biological systems and ecological interactions have more variance than can be reasonably discretely parsed. We expect that within a bayesian framework studies of pollinator behavior may be enacted via this approach at a landscape level, e.g. the scale of an entire drainage basin such as the Gunnison which is quickly becoming one of the worlds few model ecosystems. We hope that the promise of A353 probes as tools for metabarcoding play a role in these endeavors.
